Contracts for unions and intersections are hard for Nickel

### 2021-05-20 paper.tex
small fixes 5 and 2.1

For instance, we could define a flat contract \nickel{even} as
\nickel{fun x => x \% 2 == 0}, that checks
whether a number is even by checking the remainder of dividing it by 2.
And then use it to check that a function always returns an
even number, when applied to a \nickel{Integer}:

@@

\unsure{Do we need deeper examples?
Maybe something showing that flat contracts could evaluate
a function.}

@@

Even if flat contracts may seem like too permissive, the ability to make
non trivial checks on values (to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are a common pattern for safe programming, and
programmers are used to express complex predicates
over values.

@@

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let subList = fun l1 l2 =>
  lists.any (fun x => x == elt)
  && subList (list.tail l1) l2
\end{lstlisting}
  \end{center}
\caption{Inlining}
\label{fig:optimizations-inlining-ex}
\end{figure}

@@

$$M[N/x] \simeq let~x~=~N~in~M[x]$$

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to conjunctions.
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

@@

For instance, if the value \nickel{V} is contracted\todo{(Yann) Contracted means
something else I think} with
\nickel{A & (B | C)}, then, before the contract is checked,
the contract gets opened up to something equivalent to
\nickel{(A & B) | (A & C)}.

This straightforward solutions makes composing the Keil and Thiemann
system with other existing systems more difficult; in the words
of Williams, Morris, and Wadler:
"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition."\cite{RootCauseOfBlame}\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}

@@

We see two problems here, a minor one is the fact that two different
semantics need to be defined, which translates either to a difficult
to implement or a difficult to understand system\unsure{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible (as the case with common subexpression
elimination)\unsure{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}.

@@

On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus, obeys
all of their expected monitoring properties for higher ordered, union
and, intersection contracts.
However, one of these properties states the following:

@@

Which means that a continuation ($K$) complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.
Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
and dually for $K_2$, which would mean the continuation presented above would not
be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
introduction to the paper) indicates that it should.

>>>>

For instance, we could define a flat contract \nickel{even} as
\nickel{fun x => x \% 2 == 0}, that checks
whether a number is even by checking the remainder of dividing it by 2.
And then use it to check that a function always returns an
even number, when applied to an \nickel{Integer}, as shown in
Figure \ref{fig:num-to-even}.
Or even, we could come up with a contract that checks that a function
maps \nickel{0} to \nickel{0}, as can be seen on Figure
\ref{fig:0-to-0}.

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
fun f => f 0 == 0
\end{lstlisting}
    \end{center}
  \caption{A contract that checks that a function maps 0 to 0}
  \label{fig:0-to-0}
\end{figure}
\unsure{Do we need deeper examples?
Maybe something showing that flat contracts could evaluate
a function.}

@@

Even if flat contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are used to express complex predicates
over values.

@@

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let subList = fun l1 l2 =>
  lists.any (fun x => x == (lists.head l1)) l2
  && subList (list.tail l1) l2
\end{lstlisting}
  \end{center}
\caption{Inlining}
\label{fig:optimizations-inlining-ex}
\end{figure}

@@

$$M[N] \simeq let~x~=~N~in~M[x]$$

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to conjunctions.
\unsure{disjunctions, right?}
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

@@

For instance, if the value \nickel{V} is wrapped with contract
\todo{(Yann) Contracted means
something else I think.
(Teo) changed}
\nickel{A & (B | C)}, then, before the contract is checked,
it will get opened up to something equivalent to
\nickel{(A & B) | (A & C)}.

This straightforward solutions makes composing the Keil and Thiemann
system with other existing systems more difficult; in the words
of Williams, Morris, and Wadler:
"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition."\cite{RootCauseOfBlame}
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times.

@@

We see two problems here, a minor one is the fact that two different
semantics need to be defined, which translates either to a difficult
to implement or a difficult to understand system
\unsure{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) I think you're right, I'll try to reword it}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible (as the case with common subexpression
elimination)\unsure{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}.

@@

On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and, intersection contracts.
However, one of these properties states the following:

@@

Which means that a continuation ($K$) complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.

Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
and dually for $not~(f ~ true)$.
Which would mean, according to their semantics, that
the continuation presented above would not
be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
even though our intuition (and the
introduction to the paper) indicates that it should.

% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
% and dually for $K_2$, which would mean the continuation presented above would not
% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
% introduction to the paper) indicates that it should.
